---
title: "Exploratory Data Analysis"
format: 
  revealjs:
    theme: [default, inverse.scss]
---

> "Far better an approximate answer to the right question, which is often vague, than an exact answer to the wrong question, which can always be made precise."
> 
> — John Tukey

## Guiding Questions

- What does this data have to tell me about the topic?

- Is the data reliable?

- Is the data complete?

- How are the variables distributed?

- How are the variables related?


## Question Generation

- Does the structure of the data match your expectations?

- Did the data read in correctly?

- How does the data fit in context?

- Do the observations make sense? 

    - scale
    - missingness
    - relationships with other variables

## Numerical EDA

- Basic summary statistics

- Made easier with tools like `skimr` and `skimpy` that tabulate summaries and sparkline charts

```{r, include = F}
library(datasauRus)
new_data <- datasaurus_dozen
names(new_data) <- c("group", "V1", "V2")
```

```{r}
#| echo: true
#| results: false
library(skimr)
skim(new_data)
```

## Numerical EDA {.smaller}

```{r}
#| echo: false
library(skimr)
summary(skim(new_data))
```

## Numerical EDA {.smaller}

```{r}
#| echo: false
#| results: markup
library(skimr)
get_first_bit <- function(x) {
  tmp <- partition(x)[[1]]
  names(tmp)[1] <- "var"
  list(tmp[,1:3], tmp[,c(1, 4:8)])
}
res <- skim(new_data) |> get_first_bit() |> purrr::map(knitr::kable) 
res[[1]]
```
```{r}
#| echo: false
#| results: markup
res[[2]]
```


## Numerical EDA {.smaller}

```{r}
#| echo: false
#| results: markup
library(skimr)
get_sec_bit <- function(x) {
  tmp <- partition(x)[[2]]
  names(tmp)[1] <- "var"
  list(tmp[,1:5], tmp[,c(1, 6:11)])
}
res <- skim(new_data) |> get_sec_bit() |> purrr::map(knitr::kable) 
res[[1]]
res[[2]]
```
